|                                    |
|------------------------------------|
| title: Exploratory Factor Analysis |
| author: Todd Barnes                |
| date: `r Sys.Date()`               |
| output: github_document            |

```{=html}
<style type="text/css">

body{ /* Normal  */
      font-size: 12px;
  }
td {  /* Table  */
  font-size: 12px;
}
h1.title {
  font-size: 12px;
  color: DarkRed;
}
h1 { /* Header 1 */
  font-size: 12px;
  color: DarkGray;
}
h2 { /* Header 2 */
    font-size: 12px;
  color: DarkGray;
}
h3 { /* Header 3 */
  font-size: 12px;
  font-family: "Times New Roman", Times, serif;
  color: DarkGray;
}
code.r{ /* Code block */
    font-size: 10px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 12px;
}
</style>
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE )
```

```{r, include=FALSE}
setwd("C:/Users/todda/Documents/Data Repository/EFA2022")
ypix1<- read.csv("ypixefa20200515.csv",header = TRUE, sep = ",", na.strings =)
ypixdes<- read.csv("YPIXdescriptive.csv", header=TRUE,sep = ",", na.strings = "Unknown")

#library
library(psych)
library(mice)
library(GPArotation)
library(lavaan)
library(dplyr)
library(tidyr)
library(stats)
library(naniar)
library(table1)

#relabel "99" as NA if needed
ypix2<-na_if(ypix1,99)

#exclude demographic and overall variables
ypix3<- ypix1[, c(1:28)]
str(ypix3)

```

## About this project

This analysis was preceded by a literature review and a series of focus groups to determine a hypthosized set of "experiences" thought to be the most important to patients on acute psychiatric hospital units.

An initial 32 item survey was developed hypotheseizing 5 domains: Treatment effectiveness, Relationship with the Treatment Team, Nursing Presence, Unit Organization, and the Healing Environment. Te initial survey was submitted to a panel of content experts for feedback on coherence and clarity. 28-items were retained based on feedback.

The survey was piloted across five units.The results were used for quality improvement initiatives. A total of **`r nrow(ypix3)`** surveys were collected. An exploratory factor analysis was performed to understand item performance within the hypothesized domains.

## Descriptive Tables

We generate basic descriptive tables of self-reported demographic variables among our final sample. Are the respondent demographics similar to our general hospital population? *Is there significant differences of the average scores across groups?* *What are other potential sources of response bias?* *How should this impact how conclusions on internal and external validity/reliability?*

**note** A full exploratory analysis, including contingency tables is performed in separate analysis

```{r,}
table1(~Unit + Age + Gender + Race, data = ypixdes)

table1(~ Age + Gender + Race | Unit, data = ypixdes)
table1(~ Unit + Gender + Race | Age, data = ypixdes)
table1(~ Unit + Age + Race | Gender, data = ypixdes)
table1(~ Age + Gender + Unit| Race, data = ypixdes)
```

To perform our exploratory factor analysis, we limit our dataset to the 28 survey items (scored on a likert scale 1-5). Factor analysis requires numeric, non-null fields. We therefore have to remove or impute missing values. To decide if we remove or impute, we first examine for responses with a high number of missing items. Next, we analyze each question for missing values.

Responses with greater than 5% skipped questions are identified and dropped Analyze items for high missingness

```{r, collapse=TRUE}
percentmissing<- function(x){sum(is.na(x))/length(x)*100}
missing<- apply(ypix3,1,percentmissing)
replacepeople<- subset(ypix3,missing<=5)
####Total responses after excluding
nrow(replacepeople)

#once removed check to make sure columns aren't missing too much
apply(replacepeople,2,percentmissing)
nrow(replacepeople)

table(missing)

#check columns missing with original data
#all columns have less than 5% missing responses.
apply(ypix3,2,percentmissing)
```

The remaining missing responses are imputed

```{r, include=FALSE}
#impute
tempnomiss<- mice(replacepeople)
nomiss<- complete(tempnomiss,4)
summary(nomiss)
nrow(nomiss)
```

Detect and remove outliers using mahalanobis

```{r}
mahal<- mahalanobis(nomiss,
                    colMeans(nomiss),
                    cov(nomiss,use = "pairwise.complete.obs"))
plot(mahal)
hist(mahal)

summary(mahal)
boxplot(mahal)

#create the cut off score
cutoff<- qchisq(1-.001,ncol(nomiss))
ncol(nomiss)
nrow(nomiss)
summary(mahal< nomiss)

cutoff

#exclude outliers
noout<- subset(nomiss,mahal < cutoff)
nrow(noout)
```

assess correlation. Items that are too correlated should be examined for removal.

```{r}
#additivity
correl<- cor(noout, use = "pairwise.complete.obs")
symnum(correl)
```

performing a factor analysis requires us to make certain assumptions about the distribution of the data. We check our assumptions by comparing against a random variable

```{r}
#assumption set up
#make a random variable
random<- rchisq(nrow(noout),5)
fake<- lm(random~., data = noout)
standardized<- rstudent(fake)
fitted<- scale(fake$fitted.values)

#normality
hist(standardized)
#linearity
qqnorm(standardized)
abline(0,1)

#homogeneity
plot(fitted,standardized)
abline(0,0)
abline(v=0)


#correlation adequacy #yes is adequare bc p value is 0
cortest.bartlett(correl, n=nrow(noout))

#sampling adequacy #MSA >= 0.7
KMO(correl)
```

we hypothesize that these items cluster around 5 conceptual categories. We assess for statistical factors using a parallel analysis and scree plot.

``` {r{}}
#Our hypothesis was 5 factors
# parallel analysis and scree

nofactors<- fa.parallel(noout,fm="ml",fa="fa")
sum(nofactors$fa.values>1.0)
sum(nofactors$fa.values>0.7)
```

## This EFA indicates a 4 factor model. We perform EFA with oblique rotation. Results are bests formatted in excel. Items are examined for factor loadings \>0.4.

```{r,echo=FALSE}
#efa with oblique rotation, using four factor
round1<- fa(noout, nfactors =4, rotate = "oblimin", fm="ml")
round1
```

EFA source code credit to StatisticsofDoom on YT :)
